---
title: "AA502_LR_HW_2"
output: html_document
date: "2024-08-30"
---

### FIX MISSING VALUES
#### For any variable with missing values, change the data to include a missing category 
#### instead of a missing value for the categorical variable.  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#read in libraries
library(tidyverse)
library(car)

#read in training data
insurance_t_binned<-read_csv("C:\\Users\\eebla\\OneDrive\\Documents\\IAA\\AA502-Fall_I\\Homework\\LogisticRegHW\\HW2\\insurance_t_bin.csv")

#read in validation data
insurance_v_binned<-read_csv("C:\\Users\\eebla\\OneDrive\\Documents\\IAA\\AA502-Fall_I\\Homework\\LogisticRegHW\\HW2\\insurance_v_bin.csv")

#make all columns characters so all NAs can be put as Missing
insurance_t_binned<-insurance_t_binned %>%
  mutate(across(everything(), as.character))

#make all nas "missing' 
#insurance_t_binned[is.na(insurance_t_binned)] = '00 Miss' 

#check to make sure there are no more NAS, looks good
insurance_t_binned %>%
          summarise_all(~ sum(is.na(.))) %>%
          pivot_longer(everything(),names_to='variable',values_to='na_count') 
  
#make all columns in factors
insurance_t_binned<-insurance_t_binned %>%
  mutate(across(everything(), as.factor))
  
str(insurance_t_binned_to_model)
```

### SEPARATION CONCERNS
#### Check each variable for separation concerns. Document in the report and adjust any 
#### variables with complete or quasi-separation concerns. 
```{r, separation concerns}
#identify target variable
target_var<-c('INS')

#df without target variable
ins_wo_target<-insurance_t_binned %>%
  select(-c('INS'))

#mop<-as.data.frame(table(INS=insurance_t_binned$INS,insurance_t_binned$CCPURC))

#create empty dataframe to store freq counts of each variable level to the target
target_freq_df=setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("var_name","INS","level","Freq"))

#loop through each of the variables store freq of each level of predictor variables to each target variable
for (i in colnames(ins_wo_target)){
    var_name<-i
    var_df<-as.data.frame((table(INS=insurance_t_binned$INS,ins_wo_target[[var_name]])))
    names(var_df)[names(var_df) == 'Var2'] <- 'level'
    var_df$var_name=var_name
    target_freq_df<-rbind(target_freq_df,var_df)
}

#identify any 0s in the target variable frequency
#CASHBK and MMCRED are quasi-complete need to fix this
target_freq_df_zeros<-target_freq_df %>%
  filter(Freq==0)


#fix CASHBK
#look at all levels, since making a 1 would make it 50/50 anyways going to roll level 2 up to 1
table(INS=insurance_t_binned$INS,CASHBK=insurance_t_binned$CASHBK)

#there is no 2 in validation set, can't move one over
table(INS=insurance_v_binned$INS,MMCRED=insurance_v_binned$MMCRED)


#fix MMCRED
#look at #look at all levels, since making a 1 would make it 50/50 anyways going to roll 5 credits to 3 credits
#table(INS=insurance_t_binned$INS,MMCRED=insurance_t_binned$MMCRED)

#there is no level 5 in the validation dataset, can't move one over
#table(INS=insurance_v_binned$INS,MMCRED=insurance_v_binned$MMCRED)

#roll CASHBK level 3 up to 2 
#AND MMCRED level 5 to 3


insurance_t_binned_to_model<-insurance_t_binned %>%
  mutate(CASHBK=as.factor(case_when(as.integer(as.character(CASHBK))>=1 ~'1',.default=CASHBK)),
         MMCRED=as.factor(case_when(as.integer(as.character(MMCRED))>=3 ~'3',.default=MMCRED)))

#check to make sure everything is fixed, looks good.
table(INS=insurance_t_binned_to_model$INS,CASHBK=insurance_t_binned_to_model$CASHBK)
table(INS=insurance_t_binned_to_model$INS,MMCRED=insurance_t_binned_to_model$MMCRED)



#this is causing problems (I think) 

#REPORT B14,B15,B18,B19 seem to be missing CC, INV,CCPURC, INVAMOUNT 
table(BRANCH=insurance_t_binned$BRANCH,CC=insurance_t_binned$CC)
table(BRANCH=insurance_t_binned$BRANCH,INV=insurance_t_binned$INV)
table(CCPURC=insurance_t_binned$CCPURC,CC=insurance_t_binned$CC)



#look at the target with each of the BRANCHES INV,CC,CCPURC
#table(BRANCH=insurance_t_binned_to_model$BRANCH,INS=insurance_t_binned_to_model$INS)
testy<-insurance_t_binned %>% 
  filter((as.character(BRANCH)=='B15')| 
        (as.character(BRANCH)=='B14')|
        (as.character(BRANCH)=='B18')|
        (as.character(BRANCH)=='B19')) %>%
  summarise_all(~ sum(is.na(.)))


```

### GET RID OF MULTICOLLINEARITY DO NOT NEED THIS SECTION ANYMORE
```{r, include=FALSE}
#get rid of redundant variables that are causing problems
#insurance_t_binned_to_model<-insurance_t_binned_to_model%>%
#  select(-c("CC","INV"))

#alias(glm(INS ~ ., data = insurance_t_binned_to_model, family = binomial(link = "logit")))

#use VIF
#vif(full.model_after_alias)
```
### MAIN EFFECTS MODEL
#### Build a main effects only binary logistic regression model to predict the purchase of the 
#### insurance product. 
#### Use backward selection to do the variable selection – the Bank currently uses 𝛼 = 0.002 
#### and p-values to perform backward, but is open to another technique and/or significance 
#### level if documented in your report. 
#### Report the final variables from this model ranked by p-value. 
```{r, main effects model}
#MAIN EFFECTS MODEL AKA NO INTERACTIONS

#set up full and empty models to do stepwise
full.model_ins <- glm(INS ~ ., data = insurance_t_binned_to_model, family = binomial(link = "logit")) 
empty.model_ins <- glm(INS ~ 1, data = insurance_t_binned_to_model, family = binomial(link = "logit")) 

#BACKWARDS SELECTION WITH AIC, 8766.62
#INS ~ DDA + DIRDEP + NSF + IRA + LOC + INV + MTG + CC + MOVED + 
#    BRANCH + DDABAL_BIN + DEP_BIN + CHECKS_BIN + NSFAMT_BIN + 
#    TELLER_BIN + SAVBAL_BIN + ATMAMT_BIN + CDBAL_BIN + INVBAL_BIN + 
#    ILSBAL_BIN + MMBAL_BIN
#back.model_ins <- step(full.model_ins, direction = "backward",trace=FALSE)
#summary(back.model_ins)


#BACKWARDS WITH P, AIC is 8797, the backwards AIC model has a better AIC
#INS ~ DDA + NSF + IRA + INV + MTG + CC + DDABAL_BIN + 
#    CHECKS_BIN + TELLER_BIN + SAVBAL_BIN + ATMAMT_BIN + CDBAL_BIN + 
#    ILSBAL_BIN + MMBAL_BIN, family = binomial(link = "logit"), 
#    data = insurance_t_binned_to_model)
back.model_ins_p<- step(full.model_ins,direction = "backward", 
                        k = qchisq(0.002, 1, lower.tail = FALSE),trace=FALSE) 

summary(back.model_ins_p)
```

### INTERPRET COEFF AKA ODDS RATIO
#### Interpret one variable’s odds ratio from your final model as an example. 
#### Report on any interesting findings from your odds ratios from your model.
```{r,coeff interpretation}

#df with coeffs from p backwards model
coeff_first<-as.data.frame(back.model_ins_p$coefficients)
coeff_first <- tibble::rownames_to_column(coeff_first, "Variable")

#get exp of each of the coeff for odds ratio
coeff_first<-coeff_first %>%
  rename('coeff'='back.model_ins_p$coefficients') %>%
  mutate(odds_rat=exp(coeff))

#interesting things to report for odds ratios
#having a checking account DDA reduces the odds by 78% than compared to not having a checking account
# DDABAL_BIN08 > 6126.24-checking account balances above 6126.24 are 7.9360009 times more likely to buy the ins # than the reference level of checking balance (which is <=0.01)
# DDABAL_BIN07 balances >2188 and <=6126.24 are 4.2684943 times more likely to buy the ins # than the reference level of checking balance (which is <=0.01)
#DDABAL_BIN06 <= 2188.02 3.2073913
#DDABAL_BIN05 <= 1248.47 2.5779655
#DDABAL_BIN04 <= 764.81 1.9921357
#DDABAL_BIN03 <= 304.95 1.4378133
#DDABAL_BIN02 <= 82.57 1.0199862 <-doesnt really change either way good or bad
#ALL OF THE DDABAL ARE ABOVE ! THOUGH

#SAVBAL_BIN07 > 8334.97 savings balance 5.9740648 times more likely to buy the ins compared to reference level, #which is (01 <= 0.01)
#SAVBAL_BIN07 savings account (>2962.02 and <=8334.97) 3.7130475 time more likely to buy the ins compared to reference level, #which is (01 <= 0.01)
#SAVBAL_BIN06 <= 8334.97 3.7130475
#SAVBAL_BIN05 <= 2962.02 2.4305497
# SAVBAL_BIN04 <= 1259.45 1.3528059
#IN THESE TWO CASES OF SAVINGS BALANCES THERE ARE HIGHER ODDS FOR THE REFERENCE LEVEL WHICH IS <=0.01
# SAVBAL_BIN03 <= 265.87 0.7758947<- interesting it starts to have a REDUCTION IN ODDS
# SAVBAL_BIN02 <= 61.25 0.5697336<- interesting it starts to have a REDUCTION IN ODDS

#CHECKS_BIN04 > 4 0.5411992 <-reduces the odds compared to reference (01 <= 0) writing no checks
#CHECKS_BIN03 <= 4 0.9564008 (3,4)
#CHECKS_BIN02 <= 2 1.0488481 (1,2 checks)


#having an investment account is 1.83 times more likely to buy than not having an investment account



```

### LOOK FOR INTERACTIONS
#### Investigate possible interactions using forward selection including only the main effects from 
#### your previous final model. 
#### Report the final interaction variables from this model ranked by p-value.
```{r, look for interactions}
#create the full and empty model to put into forward stepwise

#lm(a ~ .^2)
#back.model_ins_p
#PUT IN ALL INTERACTIONS from the variables that came out of backwars selection above
full.model_int <- glm(INS ~ (DDA + NSF + IRA + INV + MTG + CC + DDABAL_BIN + 
    CHECKS_BIN + TELLER_BIN + SAVBAL_BIN + ATMAMT_BIN + CDBAL_BIN + 
    ILSBAL_BIN + MMBAL_BIN)^2, data = insurance_t_binned_to_model, family = binomial(link = "logit")) 
empty.model <- glm(INS ~ 1, data = insurance_t_binned_to_model, family = binomial(link = "logit")) 

#do a forward selection on all interactions within p model that we got from previous backwards selection
#using p as selection at 0.002 alpha level
for.model_int <- step(empty.model,
                  scope = list(lower = formula(empty.model),
                  upper = formula(full.model_int)),
                  direction = "forward",
                  k = qchisq(0.002, 1, lower.tail = FALSE),trace=FALSE) 

summary(for.model_int)

#interaction between IRA and DDA
final_model<-glm(formula = INS ~ SAVBAL_BIN + DDABAL_BIN + CDBAL_BIN + MMBAL_BIN + 
    INV + CHECKS_BIN + ATMAMT_BIN + TELLER_BIN + IRA + DDA + 
    CC + ILSBAL_BIN + MTG + NSF + IRA:DDA, family = binomial(link = "logit"), 
    data = insurance_t_binned_to_model)
#IRA:DDA p value is 3.129127e-04

#check for quasi seperation with the IRA,DDA interaction, looks ok, the coeff looked ok as well
insurance_t_binned_to_model %>%
  count(DDA,IRA,INS) 
```

### FINAL LOGISTIC MODEL
#### Report your final logistic regression model’s variables by significance. 
```{r,final log model}
p_values_df_final<-as.data.frame(anova(for.model_int))

#library(car)
#car::Anova(final_model, test = "LR", type = "III")

p_values_df_final <- tibble::rownames_to_column(p_values_df_final, "Variable") %>%
  filter(Variable!='NULL')


#write p values to csv
write.csv(p_values_df_final,"C:\\Users\\eebla\\OneDrive\\Documents\\IAA\\AA502-Fall_I\\Homework\\LogisticRegHW\\ps_hw_2_final.csv",row.names=FALSE)


```
