---
title: "TS_FINAL_PROJECT_ERIN"
output: html_document
date: "2024-09-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#read in libraries
library(tidyverse)
library(fpp3)
library(forecast)

#read in energy ds
energy_hw<-read_csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/energy_F2024.csv")


#convert datetime_beginning_ept from string to date and the summarize mw by day
energy_daily<-energy_hw %>% 
  mutate(date=(date(mdy_hm(datetime_beginning_ept))))%>%
  group_by(date) %>%
  dplyr::summarize(daily_mw=sum(mw))


#get the average daily usage BY MONTH
energy_monthly<-energy_daily %>% 
  mutate(date_month=(mdy(paste(month(date),'-1-',year(date)))))%>%
  group_by(date_month) %>%
  dplyr::summarize(total_average_daily_mw=mean(daily_mw))

energy_monthly<-energy_monthly %>%
  mutate(Month=yearmonth(date_month),
         month_dummy_col=relevel(as.factor(month(date_month)),ref='12'),
         oya=lag(total_average_daily_mw,n=12,order_by =date_month),
         diff_12=(total_average_daily_mw-oya)) %>%
  as_tsibble(index = Month)
 
```

### Split into train, test validate
```{r, split ds}
#split data and make them into tsibbles [1M]
#Training-August 2018 - July 2022,
#Validation-August 2022 - July 2023
#Test-August 2023 - July 2024. 
energy_train<-energy_monthly %>%
  filter(date_month<='2022-07-01') %>%
  mutate(Month=yearmonth(date_month)) %>%
  as_tsibble(index = Month)

energy_validate<-energy_monthly %>%
  filter((date_month>'2022-07-01') & (date_month<'2023-08-01')) %>%
  mutate(Month=yearmonth(date_month)) %>%
  as_tsibble(index = Month)

energy_test<-energy_monthly %>%
  filter(date_month>='2023-08-01') %>%
  mutate(Month=yearmonth(date_month)) %>%
  as_tsibble(index = Month)


```


### LAG/DIFF 12, CHEKC STATIONARITY, AUTOCORRLATION AND DO ARIMAS, CHECK RESIDUALS AFTER
## NOT ALLOWED TO USE <INSERT SAD FACE>
```{r, include=FALSE}
#p value for KPSS 0.08144748, unitroot_ndiffs is saying 0, ndiffs is saying 1
#energy_train |> features(diff_12, unitroot_kpss)
#energy_train |> features(diff_12, unitroot_ndiffs)
#ndiffs(energy_train$diff_12)

#ACF AND PACF SPIKE at 2
#ggAcf(energy_train$diff_12,lag=12)
#ggPacf(energy_train$diff_12,lag=12)


#try a bunch of ARIMA models
#energy_models_lag_12 <-energy_train %>% model(
#                    lag_ar2 = ARIMA(diff_12 ~1+ pdq(2,0,0) + PDQ(0,0,0)),
#                    lag_ma2 = ARIMA(diff_12 ~1+ pdq(0,0,2) + PDQ(0,0,0)),
#                    lag_mixed= ARIMA(diff_12~1+pdq(2,0,2) + PDQ(0,0,0)),
#                    lag_search1 = ARIMA(diff_12~1+pdq(0:12,0,0:12) +PDQ(0,0,0)),
#                    lag_searchwith_diff = ARIMA(diff_12~1+pdq(0:12,1,0:12) +PDQ(0,0,0)),
#                    lag_searchwith_diff_no_inter = ARIMA(diff_12~0+pdq(0:12,1,0:12) +PDQ(0,0,0)),
#                    lag_search2=ARIMA(diff_12~1+pdq(0:12,0,0:12)+PDQ(0,0,0),stepwise = F))
#energy_models_lag_12_df<-as.data.frame(energy_models_lag_12)
#t(energy_models_lag_12_df)
#glance(energy_models_lag_12) %>% arrange(AICc) %>% select(.model:BIC)

#lag_ar2                      ARIMA(2,0,0) w/ mean  AICc: 731.5327
#lag_ma2                      ARIMA(2,0,0) w/ mean  AICc: 731.4548	
#lag_mixed                    ARIMA(2,0,2) w/ mean  AICc: 736.1399
#lag_search1                  ARIMA(0,0,2) w/ mean  AICc: 731.4548	
#lag_searchwith_diff          ARIMA(1,1,0) w/ drift AIcc: 715.1071
#lag_searchwith_diff_no_inter ARIMA(1,1,0)          AIcc: 712.8293 <- best here
#lag_search2                  ARIMA(0,0,2) w/ mean  AICc: 731.4548	

#energy_models_lag_12 %>% select(lag_searchwith_diff_no_inter) %>% gg_tsresiduals()
#energy_models_lag_12 %>% select(lag_searchwith_diff_no_inter) %>% residuals() %>% ggPacf()
#0.5032146	 I think it is alright dof is ARIMA(1,1,0) is 1
#augment(energy_models_lag_12) %>% filter(.model=='lag_searchwith_diff_no_inter') %>% features(.innov,ljung_box, lag=12, dof = 1)

#forecast
#energy_for_lag<- energy_models_lag_12 %>% select(lag_searchwith_diff_no_inter) %>%
#fabletools::forecast(h=12)
```
### USE CLASSICAL DECOMP
```{r}
energy_class_decomp<-energy_train %>% model(classical_decomposition(total_average_daily_mw,
type = "additive"))

components(energy_class_decomp) %>% autoplot() + labs(title = "Classical additive
decomposition of Average Consumption (MW)")

season_adjusted_train<-components(energy_class_decomp) %>%
  select(Month, season_adjust,seasonal)

#p value for KPSS 0.07871977	, unitroot_ndiffs is saying 0,ndiffs saying 1, will try both
season_adjusted_train |> features(season_adjust, unitroot_kpss)
season_adjusted_train |> features(season_adjust, unitroot_ndiffs)

#ACF AND PACF SPIKE at 2
ggAcf(season_adjusted_train$season_adjust,lag=12)
ggPacf(season_adjusted_train$season_adjust,lag=12)

#try a bunch of ARIMA models
energy_models_decomp_adj <-season_adjusted_train %>% model(
                    adj_ar2 = ARIMA(season_adjust ~1+ pdq(2,0,0) + PDQ(0,0,0)),
                    adj_ma2 = ARIMA(season_adjust ~1+ pdq(0,0,2) + PDQ(0,0,0)),
                    adj_mixed= ARIMA(season_adjust~1+pdq(2,0,2) + PDQ(0,0,0)),
                    adj_ar_2_diff= ARIMA(season_adjust~1+pdq(2,1,0) + PDQ(0,0,0)),
                    adj_ma_2_diff= ARIMA(season_adjust~1+pdq(0,1,2) + PDQ(0,0,0)),
                    adj_ar_2_diff_no_int= ARIMA(season_adjust~0+pdq(2,1,0) + PDQ(0,0,0)),
                    adj_ma_2_diff_no_int= ARIMA(season_adjust~0+pdq(0,1,2) + PDQ(0,0,0)),
                    adj_search1 = ARIMA(season_adjust~1+pdq(0:12,0,0:12) +PDQ(0,0,0)),
                    adj_searchwith_diff = ARIMA(season_adjust~1+pdq(0:12,1,0:12) +PDQ(0,0,0)),
                    adj_searchwith_diff_no_inter = ARIMA(season_adjust~0+
                                                         pdq(0:12,1,0:12) +PDQ(0,0,0)),
                    adj_search2=ARIMA(season_adjust~1+pdq(0:12,0,0:12)+PDQ(0,0,0),stepwise = F))
energy_models_adj_df<-as.data.frame(energy_models_decomp_adj)
t(energy_models_adj_df)
glance(energy_models_decomp_adj) %>% arrange(AICc) %>% select(.model:BIC)


#adj_ar2                      ARIMA(2,0,0) w/ mean   AICc: 932.6079	
#adj_ma2                      ARIMA(0,0,2) w/ mean   AICc: 933.6393
#adj_mixed                    ARIMA(2,0,2) w/ mean   AICc: 937.4558	
#adj_ar_2_diff                ARIMA(2,1,0) w/ drift  AICc: 918.4713
#adj_ma_2_diff                ARIMA(0,1,2) w/ drift  AICc: 918.8667
#adj_ar_2_diff_no_int         ARIMA(2,1,0)           AICc: 916.1549	<-next best
#adj_ma_2_diff_no_int         ARIMA(0,1,2)           AICc: 916.6914	
#adj_search1                  ARIMA(0,0,2) w/ mean   AICc: 933.6393
#adj_searchwith_diff          ARIMA(1,1,1) w/ drift  AICc: 917.1612
#adj_searchwith_diff_no_inter ARIMA(1,1,1)           AICc: 914.9455  <-best here     
#adj_search2                  ARIMA(2,0,0) w/ mean   AICc: 932.6079

#adj_searchwith_diff_no_inter ARIMA(1,1,1)
energy_models_decomp_adj %>% select(adj_searchwith_diff_no_inter) %>% gg_tsresiduals()
energy_models_decomp_adj %>% select(adj_searchwith_diff_no_inter) %>% residuals() %>% ggPacf()

#0.909371 residuals check out dof is ARIMA(1,1,1)
augment(energy_models_decomp_adj) %>% filter(.model=='adj_searchwith_diff_no_inter') %>% features(.innov,ljung_box, lag=12, dof = 2)

#look at next best 
#adj_ar_2_diff_no_int  ARIMA(2,1,0)
energy_models_decomp_adj %>% select(adj_ar_2_diff_no_int) %>% gg_tsresiduals()
energy_models_decomp_adj %>% select(adj_ar_2_diff_no_int) %>% residuals() %>% ggPacf()

#0.8032039	 residuals check out dof is ARIMA(1,1,1)
augment(energy_models_decomp_adj) %>% filter(.model=='adj_ar_2_diff_no_int') %>% features(.innov,ljung_box, lag=12, dof = 2)

#forecast both
energy_for_seas_adj_1<- energy_models_decomp_adj %>% select(adj_searchwith_diff_no_inter) %>%
fabletools::forecast(h=12)

energy_for_seas_adj_2<- energy_models_decomp_adj %>% select(adj_ar_2_diff_no_int) %>%
fabletools::forecast(h=12)

#get the seasons and put them back on to get full forecasted value
season_adjusted<-season_adjusted_train %>%
  mutate(mn=month(Month)) %>%
  select(mn,seasonal) %>%
  distinct(mn,seasonal)
  
```




### MAKE REGRESSOR MATRICES FOR DUMMY AND FOURIER REGRESSORS
```{r}
#dum <- fastDummies::dummy_cols(energy_month_ts_train, select_columns = "month_dummy_col")
#month_matrix<-data.matrix((dum[,-(1:4)]))

#make matrices of month variables to regress on (to remove seasonality)
month_matrix<-model.matrix(~energy_train$month_dummy_col)
month_matrix_v<-model.matrix(~energy_validate$month_dummy_col)
month_matrix_tst<-model.matrix(~energy_test$month_dummy_col)
names<-c('intercept','Jan','Feb','Mar','April','May','June','July','Aug','Sept','Oct','Nov')

#make sure all the regressor matrices have the same column name
colnames(month_matrix)<-names
colnames(month_matrix_v)<-names
colnames(month_matrix_tst)<-names

#AICc=947.75 
#make a time series object out of the training dataset to send into the auto arima
ts_train <- ts(energy_train$total_average_daily_mw, start =c(2018,8), frequency =12)
ts_validate <- ts(energy_train$total_average_daily_mw, start = c(2022,8), frequency =12)
 
```

### ARIMAS WITH REGRESSORS (DUMMY VARIABLES) CHECK RESIDUALS AS WELL
```{r}
#Residuals from Regression with ARIMA(1,1,0) errors, ljung box p-value = 0.7697, AICc947.75 LOOKS GOOD
#ASK DR. SIMMONS ljung box dof for this is 9. 
s.reg_arima <- auto.arima(energy_train$total_average_daily_mw, xreg = month_matrix[,2:12], method="ML", seasonal = FALSE)
summary(s.reg_arima)
checkresiduals(s.reg_arima)


```

#FOURIER
```{r}
#checking what K to put in, K=2 looks the best try that since 6 is the same as dummy
plots <- list()
for (i in seq(6)) {
  fit <- auto.arima(ts_train, xreg = fourier(ts_train, K = i),
                    seasonal = FALSE, lambda = NULL)
  plots[[i]] <- autoplot(forecast::forecast(fit,
                                  xreg = fourier(ts_train, K=i, h=12))) +
    xlab(paste("K=",i,"   BIC=",round(fit$bic,2)," AICc=",round(fit$aic,2))) +
    ylab("")
}
gridExtra::grid.arrange(
  plots[[1]],plots[[2]],plots[[3]],
  plots[[4]],plots[[5]],plots[[6]], nrow=3)


#Residuals from Regression with ARIMA(1,1,1) errors, ljung box p-value = 0.9664, AICc:931.32
#K=2
four.reg_arima_2 <- auto.arima(ts_train, xreg = fourier(ts_train, K = 2), seasonal = FALSE)
summary(four.reg_arima_2)
checkresiduals(four.reg_arima_2)

#Regression with ARIMA(1,1,1) errors 
four.reg_arima_3 <- auto.arima(ts_train, xreg = fourier(ts_train, K = 3), seasonal = FALSE)
summary(four.reg_arima_3)
checkresiduals(four.reg_arima_3)


energy_models_fourier <-season_adjusted_train %>% model(
                    fourier_k3_search1 = ARIMA(season_adjust~1+pdq(0:12,0,0:12) 
                                            +PDQ(0,0,0)+fourier(K=3),stepwise=F),
                    fourier_k3_search_2 = ARIMA(season_adjust~0+pdq(0:12,1,0:12) +PDQ(0,0,0)+
                                               fourier(K=3),stepwise=F),
                    fourier_k3_search_3 = ARIMA(season_adjust~1+pdq(0:12,1,0:12) +PDQ(0,0,0)+
                                               fourier(K=3),stepwise=F),
                    fourier_k2_search1 = ARIMA(season_adjust~1+pdq(0:12,0,0:12) 
                                            +PDQ(0,0,0)+fourier(K=2),stepwise=F),
                    fourier_k2_search_2 = ARIMA(season_adjust~0+pdq(0:12,1,0:12) +PDQ(0,0,0)+
                                               fourier(K=2),stepwise=F),
                    fourier_k2_search_3 = ARIMA(season_adjust~1+pdq(0:12,1,0:12) +PDQ(0,0,0)+
                                               fourier(K=2),stepwise=F))
energy_models_fourier_df<-as.data.frame(energy_models_fourier)
t(energy_models_fourier_df)
glance(energy_models_fourier) %>% arrange(AICc) %>% select(.model:BIC)

#WITH STEPWISE TRUE
#fourier_k3_search1  LM w/ ARIMA(0,0,2) errors 949.8826	
#fourier_k3_search_2 LM w/ ARIMA(1,1,1) errors 930.4812<-best of K3 autoarima chose same
#fourier_k3_search_3 LM w/ ARIMA(1,1,1) errors 933.5593	

#fourier_k2_search1  LM w/ ARIMA(0,0,2) errors 944.1589	
#fourier_k2_search_2 LM w/ ARIMA(1,1,1) errors 924.9261 <-best of K2 same as below auto.arima chose same
#fourier_k2_search_3 LM w/ ARIMA(1,1,1) errors 927.6707	

#WITH STEPWISE F
#fourier_k3_search1  LM w/ ARIMA(2,0,0) errors 948.7736	
#fourier_k3_search_2 LM w/ ARIMA(1,1,1) errors 930.4812	<-best same as above
#fourier_k3_search_3 LM w/ ARIMA(1,1,1) errors 933.5593 

#fourier_k2_search1  LM w/ ARIMA(2,0,0) errors 943.0443	
#fourier_k2_search_2 LM w/ ARIMA(1,1,1) errors 924.9261 <-best, same as above auto.arima chose same
#fourier_k2_search_3 LM w/ ARIMA(1,1,1) errors 927.6707

```

### FORECAST WITH THE AUTO ARIMAS CHOSEN ABOVE
```{r}
#forecast for the dummy variable regressors, using validation regressor matrix, automatically will pick the horizon
sd_for<-forecast::forecast(s.reg_arima, xreg=month_matrix_v[,2:12])
#plot
autoplot(forecast::forecast(s.reg_arima, xreg=month_matrix_v[,2:12])) + 
autolayer(fitted(s.reg_arima), series="Fitted") + ylab("MW")


#forecast for fourier regressors #K=2
four_for<-forecast::forecast(four.reg_arima_2, xreg=fourier(ts_validate, K = 2,h=12))
#plot
autoplot(forecast::forecast(four.reg_arima_2, xreg=fourier(ts_validate, K = 2,h=12))) + 
autolayer(fitted(four.reg_arima_2), series="Fitted") + ylab("MW")


four_for_3<-forecast::forecast(four.reg_arima_3, xreg=fourier(ts_validate, K = 3,h=12))
#plot
autoplot(forecast::forecast(four.reg_arima_3, xreg=fourier(ts_validate, K = 3,h=12))) + 
autolayer(fitted(four.reg_arima_3), series="Fitted") + 
  labs(x='Month',y='Average Daily Consumption (MW)',
       title='Fourier K=3 Forecast for Daily Consumption (MW)')
```


### PUT ALL FORECASTS TOGETHER AND GET ACCURACY METRICS
```{r}
#put all forecasts on validation set to check with others REGRESSION WITH DUMMY AND FOURIER 6 ARE SAME
energy_validate_final<-energy_validate %>% dplyr::bind_cols(sd_for=c(sd_for$mean),four_2=c(four_for$mean),
                              four_3=c(four_for_3$mean),
                              for_decomp1 =c(energy_for_seas_adj_1$.mean),
                              for_decomp2=c(energy_for_seas_adj_2$.mean)) %>%
  mutate(mn=as.integer(as.character(month_dummy_col))) %>%
  left_join(season_adjusted) %>%
  mutate(forecast_season_added_decomp1=seasonal+for_decomp1,
         forecast_season_added_decomp2=seasonal+for_decomp2)


#seasonal dummy MAPE and MAE (same as fourier 6)
#MAE is 5756.52
#MAPE is 5.52% THIS IS WORSE THAN THE HOLT WINTERS ESMs
sd.error<-energy_validate_final$total_average_daily_mw - energy_validate_final$sd_for
sd.MAE <- mean(abs(sd.error))
sd.MAPE <- mean(abs(sd.error)/abs(energy_validate_final$total_average_daily_mw))*100

#fourier,K=2 MAPE AND MAE
#MAPE is 5.216%, MAE is 5374.12 
f2.error<-energy_validate_final$total_average_daily_mw - energy_validate_final$four_2
f2.MAE <- mean(abs(f2.error))
f2.MAPE <- mean(abs(f2.error)/abs(energy_validate_final$total_average_daily_mw))*100


#fourier, K=3 MAPE AND MAE
#MAPE is 5.206%, MAE is 5363.12 THIS ONE IS THE BEST  
f3.error<-energy_validate_final$total_average_daily_mw - energy_validate_final$four_3
f3.MAE <- mean(abs(f3.error))
f3.MAPE <- mean(abs(f3.error)/abs(energy_validate_final$total_average_daily_mw))*100

#seasonal decomp MAPE AND MAE
#MAPE is 5.695%, MAE is 5958.02
decomp_1.error<-energy_validate_final$total_average_daily_mw - energy_validate_final$forecast_season_added_decomp1
decomp_1.MAE <- mean(abs(decomp_1.error))
decomp_1.MAPE <- mean(abs(decomp_1.error)/abs(energy_validate_final$total_average_daily_mw))*100

#seasonal decomp MAPE AND MAE
#MAPE is 5.705%, MAE is 5972.05 
decomp_2.error<-energy_validate_final$total_average_daily_mw - energy_validate_final$forecast_season_added_decomp2
decomp_2.MAE <- mean(abs(decomp_2.error))
decomp_2.MAPE <- mean(abs(decomp_2.error)/abs(energy_validate_final$total_average_daily_mw))*100
```

### GRAPH THEM ALL TOGETHER
```{r}
#graph them all together
#fourier, K=3 MAPE is 5.206%, MAE is 5363.12 THIS ONE IS THE BEST  
ts_models_all<-energy_validate_final %>% ggplot(aes(x=Month))+
  geom_line(aes(y=total_average_daily_mw,color='Original Series'),size=0.5)+
  geom_line(aes(y=sd_for,color="Seasonal Dummy"),size=0.5)+
  geom_line(aes(y=four_3,color="Fourier K=3"),size=0.5)+
  geom_line(aes(y=four_2,color="Fourier K=2"),size=0.5)+
  geom_line(aes(y=forecast_season_added_decomp1,color="Decomp 1 Model"),size=0.5)+
  geom_line(aes(y=forecast_season_added_decomp2,color="Decomp 2 Model"),size=0.5)+
  labs(x='Month',y='Average Daily Consumption (MW)',title='Forecasts Vs. Validation',color='Models')+
  scale_color_manual(values = c("Original Series"='grey',
                                "Seasonal Dummy" = "cornflowerblue", 
                                "Fourier K=3" = "#D55E00",
                                "Fourier K=2"='purple',
                                "Decomp 1 Model"='green',
                                "Decomp 2 Model"='red')) +
  theme_classic()+
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",) +
  scale_x_yearmonth(date_breaks = "6 months", date_labels = "%b %Y")


ggsave(ts_models_all, 
       filename = "ts_models_all.jpg",
       device = "jpg",
       height = 6, width = 12, units = "in")

```
