---
title: "AA502_TS_HW2_Erin"
output: html_document
date: "2024-08-30"
---
## Objective
#### PJM is a regional transmission organization (RTO) that provides resources for the market of
#### wholesale electricity in all or parts of 13 states and the District of Columbia (source: PJM.com).
#### PJM, henceforth referred to as the client, would like to contract your services in understanding
#### metered load Megawatts for the AEP Appalachian Power transmission zone. For this analysis,
#### the client would like the data to be rolled up to total daily MW and averaged by month. The
#### training data should be from August 2018-July 2022; the validation data set should be from
#### August 2022-July 2023; and the test data should be from August 2023-July 2024. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#read in libraries
library(tidyverse)

#read in energy ds
energy_hw<-read_csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/energy_F2024.csv")


#convert datetime_beginning_ept from string to date and the summarize mw by day
energy_daily<-energy_hw %>% 
  mutate(date=(date(mdy_hm(datetime_beginning_ept))))%>%
  group_by(date) %>%
  dplyr::summarize(daily_mw=sum(mw))


#get the average daily usage BY MONTH
energy_monthly<-energy_daily %>% 
  mutate(date_month=(mdy(paste(month(date),'-1-',year(date)))))%>%
  group_by(date_month) %>%
  dplyr::summarize(total_average_daily_mw=mean(daily_mw))


energy_monthly_t.s<-energy_monthly %>%
  mutate(Month=yearmonth(date_month)) %>%
  as_tsibble(index = Month)
```

### Split into train, test validate
```{r, split ds}
#Training-August 2018 - July 2022,
#Validation-August 2022 - July 2023
#Test-August 2023 - July 2024. 
energy_train<-energy_monthly %>%
  filter(date_month<='2022-07-01')
summary(energy_train)

energy_validate<-energy_monthly %>%
  filter((date_month>'2022-07-01') & (date_month<'2023-08-01'))

energy_test<-energy_monthly %>%
  filter(date_month>='2023-08-01')

energy_monthly<-energy_monthly %>%
  mutate(color_type=case_when(date_month<='2022-07-01'~'Training',
                                       ((date_month>'2022-07-01') &(date_month<'2023-08-01')) ~'Validation',
                                        date_month>='2023-08-01'~'Test'))

#shows the entire dataset by avg daily mw usage wiht data splits
splits<-energy_monthly%>%
  ggplot(aes(date_month,total_average_daily_mw,color=color_type))+
  geom_path(size=1.2)+
  scale_color_manual(name="Splits",values = c("Training" = "black","Validation"='cornflowerblue', 
                     "Test" = "#D55E00")) +
  labs(x='Month',y='Average Daily Consumption (MW)',title='Average Daily Consumption (MW) Over Time With Splits')+
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right" ) +
  scale_x_yearmonth(date_breaks = "6 months", date_labels = "%b %Y")


ggsave(splits, 
       filename = "splits.jpg",
       device = "jpg",
       height = 6, width = 12, units = "in")

```

### PLOT
#### Visual of the training data set overlaid with the trend (be sure to indicate what
#### type of decomposition was used).
```{r}
#make energy train into a tsibble, have to make sure the month is in the form tsibble wants
energy_month_ts_train <- energy_train %>%
  mutate(Month=yearmonth(date_month)) %>%
  as_tsibble(index = Month)

#check that is read in correct
energy_month_ts_train
  
energy_month_ts_validate <- energy_validate %>%
  mutate(Month=yearmonth(date_month)) %>%
  as_tsibble(index = Month)

energy_month_ts_test<- energy_test %>%
  mutate(Month=yearmonth(date_month)) %>%
  as_tsibble(index = Month)

energy_time_plot<-energy_month_ts_train %>% autoplot(colour='blue')+
 labs(y ="Total Avg Daily MW", x='Month',title = "Total Average Daily MW By Month" )

ggsave(energy_time_plot, 
       filename = "energy_time_plot.jpg",
       device = "jpg",
       height = 7, width = 9, units = "in")


#STL decomp, and plot it
energy_decomp <- energy_month_ts_train %>% model(stl = STL(total_average_daily_mw))

full_decomp<-components(energy_decomp) %>% 
  autoplot() + 
  theme_classic()+
  labs(y ="Average Daily Consumption (MW)", x='Month',title = "STl Decomposition of Average Daily Consumption (MW)" )

ggsave(full_decomp, 
       filename = "full_decomp.jpg",
       device = "jpg",
       height = 5, width = 12, units = "in")

#overlay trend on dataset in a graph
trend_overlay<-components(energy_decomp) %>% as_tsibble() %>%
autoplot(total_average_daily_mw, colour="black") +
geom_line(aes(y=trend), colour = "#D55E00") + 
labs( y ="Average Daily Consumption (MW)", x='Month',title = "Average Daily Consumption (MW) With Trend Overlaid" )+
theme_classic()

ggsave(trend_overlay, 
       filename = "trend_overlay.jpg",
       device = "jpg",
       height = 6, width = 12, units = "in")


#lets try diff decomp, 

#ARIMA DECOMP- can handle outliers
#library(seasonal)


#energy.tsob_train<-ts(energy.tsob_train,start=c(2018,8),frequency=12)

#decom <- seas(energy.tsob_train, x11 = "")

#x11_dcmp_t <- energy_month_ts_train |>
#  model(x11 = X_13ARIMA_SEATS(total_average_daily_mw ~ x11())) %>%
#  components()


#graph
#autoplot(x11_energy_dcmp)+ 
#labs(title = "Decomposition of Average Daily MW USE With X-11.")

#decomposed seasonality only 
decomp_season_graph<-components(energy_decomp) %>% as_tsibble() %>%
ggplot(aes(x=Month,y=season_year), colour="cornflowerblue")+
geom_line(size=1.2,colour="cornflowerblue")+
labs( y ="Average Daily Consumption (MW)", x='Month',title = "Seasonality Component of STL Decomposition" )+
theme_classic()+
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right" ) +
  scale_x_yearmonth(date_breaks = "6 months", date_labels = "%b %Y")
  

ggsave(decomp_season_graph, 
       filename = "decomp_season_graph.jpg",
       device = "jpg",
       height = 6, width = 12, units = "in")


#overlay seasonal on dataset in a graph
seasonal_adj_w_trend<-components(energy_decomp) %>% as_tsibble() %>% 
  ggplot(aes(x=Month))+
  geom_line(aes(y=total_average_daily_mw,color='Original Series'),size=1) + 
  geom_line(aes(y=season_adjust, color ="Seasonal Adjusted"),size=1.2)+
  geom_line(aes(y=trend,color="Trend Only"),size=1.2) + 
  scale_color_manual(values = c("Original Series"='grey',
                                "Seasonal Adjusted" = "cornflowerblue", 
                                "Trend Only" = "#D55E00")) +
  labs(y = "Average Daily Consumption (MW)",
       x='Month', 
       title = "Seasonally Adjusted Average Daily Consumption (MW)",
       color='Components')+
  theme_classic()+
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  scale_x_yearmonth(date_breaks = "6 months", date_labels = "%b %Y")
  
ggsave(seasonal_adj_w_trend, 
       filename = "seasonal_adj_w_trend.jpg",
       device = "jpg",
       height = 6, width = 12, units = "in")


#autoplot(season_year, colour="black") +
#labs( y ="Total Avg Daily MW", title = "Total Avg Daily MW Season year" )+
#scale_x_yearmonth(date_breaks = "4 month", date_labels = "%m-%Y")+
#theme(axis.text.x = element_text(angle = 90, hjust = 1))

#strength of trend/seasonality
energy_month_ts_train %>% features(total_average_daily_mw,feat_stl)

```

### ESM target is average daily MW per Month
```{r, }
#TEST ALL THE MODELS
#From decomp, seasonality is present, trend is not as present

#MAPES from belows test
#HWAAA- 3.788538	
#HWMultMAM-3.715035	 <-----------BEST IS HW multiplicative seasonal and Additive trend
#HWauto	Test-5.199704
#HWdAAdA-4.539581
#HWdANA	-5.202978
#HWdMAdM-4.364705
#HWdMNM-4.990485
#fit many different ESM models
energy_fit <- energy_month_ts_train %>% 
model(HWAAA = ETS(total_average_daily_mw ~ error("A") + trend("A") + season("A")),    
      HWMultMAM = ETS(total_average_daily_mw ~ error("M") + trend("A") + season("M")),    
      HWdMAdM = ETS(total_average_daily_mw ~ error("M") + trend("Ad") + season("M")),
      HWdAAdA = ETS(total_average_daily_mw ~ error("A") + trend("Ad") + season("A")),
      HWdANA = ETS(total_average_daily_mw ~ error("A") + trend("N") + season("A")),
      HWdMNM = ETS(total_average_daily_mw ~ error("M") + trend("N") + season("M")),
      HWauto= ETS(total_average_daily_mw))

#forecast the mw, HWMult has the best MAPE
energy_fc <- energy_fit %>% fabletools::forecast(h = 12)
fabletools::accuracy(energy_fc, energy_month_ts_validate)


```


## PLOT BEST MODEL
```{r, plot best model}
#BEST MODEL (MAM)
#multiplicative seasonality, MAPE for the validation set is 3.715035%
MAM.energy<- energy_month_ts_train %>%  model(ETS(total_average_daily_mw ~ error("M") + 
trend("A") + season("M")))
energy.for <- MAM.energy %>%  fabletools::forecast(h = 12)

#plot HWmult without confidence intervals, need to get legends for the lines
forecast_plot<-energy.for %>% autoplot(energy_month_ts_validate, level = NULL) +  
  geom_line(aes(y = .fitted,color='Original Series')
            data = augment(MAM.energy),size=1.2) + 
  geom_line(aes(y=total_average_daily_mw),size=1.2,col='#D55E00')
  labs(y="Average Daily Consumption (MW)", 
       x='Month',
       title="TAverage Daily Consumption (MW) With Forecast") +
  theme_classic()+
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  scale_x_yearmonth(date_breaks = "6 months", date_labels = "%b %Y")


  ggplot(aes(x=Month))+
  geom_line(aes(y=log_mw,color='Original Series (Log Scale)'),size=1) + 
  geom_line(aes(y=season_adjust, color ="Seasonal Adjusted"),size=1.2)+
  geom_line(aes(y=trend,color="Trend"),size=1.2) + 
  scale_color_manual(values = c("Original Series (Log Scale)"='grey',
                                "Seasonal Adjusted" = "cornflowerblue", 
                                "Trend" = "#D55E00")) +

ggsave(forecast_plot, 
       filename = "forecast_plot.jpg",
       device = "jpg",
       height = 6, width = 12, units = "in")

```




### PLaying around
```{r, playing around}
#CHECK ANOTHER (ANA)
ANA.energy<- energy_month_ts_train %>%  model(ETS(total_average_daily_mw ~ error("A") + 
trend("N") + season("A")))
energy.for.ANA <- ANA.energy %>%  fabletools::forecast(h = 12)

#plot HWmult
energy.for.ANA %>% autoplot(energy_monthly_t.s) +  geom_line(aes(y = .fitted), 
col="#D55E00", data = augment(ANA.energy)) +  
labs(y="Avg Daily MW Per Month", title="Total MW Usage Over Time") +  
guides(colour = "none")


#CHECK ANOTHER (MNM)
MNM.energy<- energy_month_ts_train %>%  model(ETS(total_average_daily_mw ~ error("M") + 
trend("N") + season("M")))
energy.for.MNM <- MNM.energy %>%  fabletools::forecast(h = 12)

#plot HWmult
energy.for.MNM %>% autoplot(energy_monthly_t.s) +  geom_line(aes(y = .fitted), 
col="#D55E00", data = augment(MNM.energy)) +  
labs(y="Avg Daily MW Per Month", title="Total MW Usage Over Time") +  
guides(colour = "none")

```

